Progress Report on NHANES Data

### 2023-10-03 Docs/ProgressReportAppl.Rmd
explore projection of eigenfunctions

---
title: "Progress Report"
author: "Ying Jin"
date: "2023-09-15"
output: 
  html_document:
    self_contained: no
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(516)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
theme_set(theme_minimal())

# df <- read_rds(here("Data/nhanes_bi.rds"))
# load(here("Data/ApplOutput_fGFPCA.RData"))

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))

# Simulation

I am now on the way to add the debias step to the whole algorithm. Just so that I have a set of "true parameters" to comparision, I am gonna use the simulation dataset for this part.

## Generation mechanism

- I am gonna generate only one dataset as an example
- 500 individuals each with 1000 measures

\[\begin{aligned}
Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}\]

where:

- $t$ is equal-spaced on $[0, 1]$
- $f_0(t)=0$
- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively.

```{r gen_data, class.source='show'}
N <- 500 # sample size
J <- 1000 # number of observation points

t = seq(0,1,len=J) # observations points

# mean function
f_0 <- function(s) 0

#eigenfunctions 
K <- 4 # number of eigenfunctions
phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
                     sin(4*pi*t),cos(4*pi*t))

# eigenvalues
lambda = 0.5^(0:(K-1))

# generated data 
## score
xi <- matrix(rnorm(N*K),N,K)
xi <- xi %*% diag(sqrt(lambda))

# subject-specific random effect
b_i <- xi %*% t(phi); # of size N by J

# latent gaussian function
eta_i <- t(vapply(1:N, function(x){
    f_0(t) + b_i[x,]
  }, numeric(J)))

# outcome binary function
Y_i <- matrix(rbinom(N*J, size=1, prob=plogis(eta_i)), 
                N, J, byrow=FALSE)

# format into dataframe
# id = subject identifier (factor variable)
# sind = numeric value corresponding to the observed functional domain
# Y = functional response (binary)
# sind_inx = numeric value associated with order of "sind"
#            this is not necessary, but may be of convenience when you implement the method
df <- data.frame(id = factor(rep(1:N, each=J)),
                 t = rep(t, N), 
                 Y = as.vector(t(Y_i)),
                 eta_i = as.vector(t(eta_i)),
                 sind = rep(1:J, N))
# visualization
rand_id <- sample(N, size = 4)
df %>% filter(id %in% rand_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y))+
  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
  facet_wrap(~id)
```

## fGFPCA model fitting

### Bin ata

- Bin every 10 consecutive observations

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
# unique(df$bin)

df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~id)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```

## Data split

- 400 subjects for training, 100 for out-of-sample prediction

```{r}
train_id <- sample(unique(df$id), size = N*0.8)
test_id <- setdiff(unique(df$id), train_id)

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
```

### Local GLMM

- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
- Fit on the training set

```{r, cache=TRUE}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

# local GLMM and estimate latent function
# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
t_local_glmm <- t2-t1 
```

```{r}
df_est_latent <- bind_rows(df_est_latent) 
# head(df_est_latent)

# example estimated latent function
df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  mutate(eta_i = exp(eta_i)/(1+exp(eta_i))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_line(aes(x=sind, y=eta_i, group = id, col = "true"))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

### FPCA

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)

mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time
# dim(mat_est_unique)

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
```

```{r}
plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
```

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2,2))

plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

```{r}
# plot correlation matrix
heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
```

### Debias (step 4)

According to the fGFPCA paper, in this step we do two things:

1. Project the eigenfunctions with B-spline basis back to the original grid
2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction

#### Debias with GLMM model

First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation

```{r}
df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
colnames(df_phi) <- c("bin", paste0("phi", 1:4))

# train_df$bin <- as.numeric(as.character(train_df$bin))
train_df <- train_df %>% left_join(df_phi, by = "bin")
train_df$id <- as.factor(train_df$id)
```

- Here, what if I replace "bin" with "t"?

```{r, class.source = "fold-show"}
# usethis::edit_r_environ()
t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, data=train_df %>% filter(bin==sind), 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t_debias <- t2-t1
```

- And now I try to extract the variance of scores (debiased eigenvalues). 
- Use the conclusion that connects smoothing parameters to variance

$$\hat{p}_k=1/\hat{\lambda}_k^2$$

```{r}
fpca_mod$evalues
new_lambda <- 1/debias_glmm$sp[2:5]
new_lambda/lambda
fpca_mod$evalues[1:4]/new_lambda
```

- Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.

#### Projection

Then I would like to project eigenfunctions using B-spline basis functions:

- Would you interpolate mean and eigenfunctions separately (in my case, five interpolations)? 
- I did not find the reeval_efunctions function

```{r, fig.height=5, fig.width=5, class.source = "fold-show"}
# mean
interp_mean <- interpSpline(mid, fpca_mod$mu)
full_mean <- predict(interp_mean, x=1:J)
plot(mid, fpca_mod$mu, ylab = "mean", pch = 20)
lines(full_mean$x, full_mean$y, col = "blue")

# PCs
interp_phi1 <- interpSpline(mid, fpca_mod$efunctions[, 1])
full_phi1 <- predict(interp_phi1, x=1:J)
plot(mid, fpca_mod$efunctions[, 1], ylab = "pc1", pch=20)
lines(full_phi1$x, full_phi1$y, col = "blue")

interp_phi2 <- interpSpline(mid, fpca_mod$efunctions[, 2])
full_phi2 <- predict(interp_phi2, x=1:J)
plot(mid, fpca_mod$efunctions[, 2], ylab = "pc2", pch=20)
lines(full_phi2$x, full_phi2$y, col = "blue")

interp_phi3 <- interpSpline(mid, fpca_mod$efunctions[, 3])
full_phi3 <- predict(interp_phi3, x=1:J)
plot(mid, fpca_mod$efunctions[, 3], ylab = "pc3", pch=20)
lines(full_phi3$x, full_phi3$y, col = "blue")

interp_phi4 <- interpSpline(mid, fpca_mod$efunctions[, 4])
full_phi4 <- predict(interp_phi4, x=1:J)
plot(mid, fpca_mod$efunctions[, 4], ylab = "pc4", pch=20)
lines(full_phi4$x, full_phi4$y, col = "blue")
```

Perhaps I should try with the simulation data.

>>>>

---
title: "Progress Report on NHANES Data"
author: "Ying Jin"
date: "2023-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
